---
title: "pangoLEARN vs. Usher Classifications using pangolin"
date: "Updated:  `r format(Sys.time(), '%B %d, %Y')`<div style=\"height: 40px;\"></div>"
output:
  html_document:
    toc: true
    toc_depth: 5
---

```{r echo = F, warning = F, message = F}
library(dplyr)
library(DT)
library(stringr)

plearn <- list.files("/NetDrive/Projects/COVID-19/Other/master_fasta/", pattern = "*plearn*")
lineages_plearn <- read.csv(paste("/NetDrive/Projects/COVID-19/Other/master_fasta/", plearn,
                                  sep=""), header = TRUE, na.strings=c("","NA"), stringsAsFactors=FALSE)

usher <- list.files("/NetDrive/Projects/COVID-19/Other/master_fasta/", pattern = "*usher*")
lineages_usher <- read.csv(paste("/NetDrive/Projects/COVID-19/Other/master_fasta/", usher,
                                  sep=""), header = TRUE, na.strings=c("","NA"), stringsAsFactors=FALSE)

lineages_plearn_sub <- lineages_plearn %>% select(taxon, lineage) %>% rename(plearn_lin = lineage)

lineages_usher_sub <- lineages_usher %>% select(taxon, lineage) %>% rename(usher_lin = lineage)

merged <- merge(lineages_plearn_sub, lineages_usher_sub, by = "taxon")
```

## Overview

The comparisons below highlights the assignment differences and similarities between using:
\
\
1. a combination of pangoLEARN and scorpio
\
2. a combination of scorpio and Usher
\
\
for lineage assignments using pangolin. These differences are shown across identical versions of pangolin and pango designations. 
\

The motivation to compare the two methods of assignment was suggested by the Rambaut group as method for validating potential false positive lineage calls with pangoLEARN. Specifically, any lineage assignments for Delta sub-lineages from pangoLEARN were suggested to be validated by Usher in an effort to reduce the false positive rate on AY lineage calling with pangoLEARN. This was first reported for pangoLEARN 2021-08-09 and attempts were made to correct these assignments in the current version: `r format (unique(lineages_plearn$pangoLEARN_version)[1])`. 
\
\
Full version details of all modules used in Pangolin can be found below:

\
\


## Version information


pangolin: **`r format (unique(lineages_plearn$pangolin_version)[1])`**

pangoLEARN: **`r format (unique(lineages_plearn$pangoLEARN_version)[1])`**

pango designation: **`r format (unique(lineages_plearn$pango_version)[1])`**

Usher: **`r format (unique(lineages_usher$version)[2])`**

\
\

## AY & AZ lineages reported by pangoLEARN and Usher

### pangoLEARN

```{r, echo = F, message = F, warning = F}

datatable(lineages_plearn_sub %>% group_by(plearn_lin) %>% summarise(Counts = n()) %>%
  subset(grepl("AY|AZ", plearn_lin)) %>% filter(Counts > 0) %>% arrange(-Counts) %>%
  rename(Lineage = plearn_lin),
           rownames = FALSE,
           escape=FALSE,
          class="compact",
          width = 800)

```
### Usher

```{r, echo = F, message = F, warning = F}

datatable(lineages_usher_sub %>% group_by(usher_lin) %>% summarise(Counts = n()) %>%
  subset(grepl("AY|AZ", usher_lin)) %>% filter(Counts > 0) %>% arrange(-Counts) %>%
  rename(Lineage = usher_lin),
           rownames = FALSE,
           escape=FALSE,
          class="compact",
          width = 800)

```

## Assignment differences, pangoLEARN and Usher, AY & AZ lineages
\
A summary of the tallies of different lineage assignments obtained through either pangoLEARN or Usher can be seen below for any AY (Delta sub-lineage) and AZ (B.1.1.318 sub-lineage)

\
\

```{r, echo = F, message = F, warning = F}
differences <- subset(merged, plearn_lin != usher_lin)
difference_combos <- as.data.frame(table(differences$plearn_lin, differences$usher_lin)) %>% filter(Freq > 0)
colnames(difference_combos) <- c("pangoLEARN lineage", "Usher lineage", "Frequency")
difference_combos <- difference_combos[with(difference_combos, grepl("AY|AZ", paste(`pangoLEARN lineage`, `Usher lineage`))),]

all_plearn <- lineages_plearn_sub %>% group_by(plearn_lin) %>% summarise(counts = n())

difference_combos <- merge(difference_combos, all_plearn, by.x = "pangoLEARN lineage",
                           by.y = "plearn_lin") %>% 
  mutate(`percent of all Possible*` = round(100*(Frequency/counts), digits = 1)) %>% select(-counts) %>%
  arrange(-Frequency)

datatable(difference_combos,
           rownames = FALSE,
           escape=FALSE,
          class="compact",
          width = 800)

```

<br>
<br>


## Identical assignments, pangoLEARN and Usher, AY & AZ lineages
\
A summary of the tallies of the same lineage assignment obtained through either pangoLEARN or Usher can be seen below for any AY (Delta sub-lineage) and AZ (B.1.1.318 sub-lineage)
\
\
```{r, echo = F, message = F, warning = F}
same_ay_az <- subset(merged, plearn_lin == usher_lin & grepl("AY|AZ", plearn_lin))
same_ay_az <- as.data.frame(table(same_ay_az$plearn_lin, same_ay_az$usher_lin)) %>% filter(Freq > 0) %>% arrange(-Freq)
colnames(same_ay_az) <- c("pangoLEARN lineage", "Usher lineage", "Frequency")

same_ay_az <- merge(same_ay_az, all_plearn, by.x = "pangoLEARN lineage",
                           by.y = "plearn_lin") %>% 
  mutate(`percent of all Possible*` = round(100*(Frequency/counts), digits = 1)) %>% select(-counts) %>%
  arrange(-Frequency)

datatable(same_ay_az,
           rownames = FALSE,
           escape=FALSE,
          class="compact",
          width = 800)

```

<br>

* Note: percent of all Possible is measured by the total number of sequences using pangoLEARN that have the particular lineage assignment. For example, a percentage of 100% as shown in the differences chart would indicate that 100% of sequences assigned to that particular lineage in pangoLEARN were not supported by Usher, and vice versa if displayed in the identical chart. 
\
\

## Defining mutation presence for AY lineages

\

The table below highlights the patterns of concordance and discordance among AY lineages when detecting a particular defining mutation. For example, a lineage with a percentage of 100% with the defining mutation indicates that all of the sequences designated as the particular AY lineage were found to have the defining mutation corresponding to that specific AY number, likely indicating a true positive. Conversely, a percentage of sequences without the defining mutation indicates that the mutation was not seen, potentially indicating a false positive. The percentages and total counts and shown for both Usher and pangoLEARN. 

If either of Usher or pangoLEARN outputs an AY lineage where more than 10% of the sequences are found to not have the defining mutation, they are flagged in the final column. 
\
\

```{r, echo = F, message = F, warning = F}

find_nextclade <- list.files("/NetDrive/Projects/COVID-19/Other/master_fasta/", pattern = "nextclade_all*")
nextclade_all <- read.csv(paste("/NetDrive/Projects/COVID-19/Other/master_fasta/", find_nextclade, sep=""), header = TRUE, na.strings=c("","NA"), stringsAsFactors=FALSE,
                          sep="\t")

```

```{r, echo = F, message = F, warning = F}

ay_mutations <- read.csv("ay_defining_mutations.tsv", sep = "\t", header = TRUE, na.strings=c("","NA"), stringsAsFactors=FALSE)

ay_vec <- ay_mutations$lineage
names(ay_vec) <- ay_mutations$defining_mutation

searches <- data.frame(lineage = as.character(),
                       search_term = as.character(), stringsAsFactors = F)

for (elem in ay_vec) {
  # split the strings into the individual mutations 
  string_split <- as.vector(str_split(names(which(ay_vec == elem)), ";", 5))
  
  if (length(string_split[[1]]) > 1) {
    # if there is more than one defining mutation, find all pairwise combos and create grep string
    # for each one
    combinations <- combn(string_split[[1]], 2)
    search <- vector()
    for(i in 1:ncol(combinations)) {      
    unlisted <- unlist(combinations[,i])
    term <- paste(unlisted[1], ".*", unlisted[2], "|", unlisted[1], ".*", unlisted[2], sep="")
    # append all the pairwise grep searches to a vector
    search <- c(search, term)
    }
    
    # collapse the grep searches if they are longer than one term
    
    term_merged <- paste(search, collapse = "|")

    frame_multiple <- data.frame(lineage = elem,
                                 search_term = term_merged)
    
    # add the searches to the data frame, either single of multiple mutations
    searches <- rbind(searches, frame_multiple)
  } else {
    frame_single <- data.frame(lineage = elem,
                               search_term = string_split[[1]])
    
    searches <- rbind(searches, frame_single)
  }
}
  

```


```{r, echo = F, message = F, warning = F}

# append total counts and individual module percentages to a frame to render
# in table
ay_tracking_frame <- data.frame(lineage = as.character(), num_plearn = as.numeric(),
                                defining_mut = as.character(),
                                perc_plearn_with_mut = as.numeric(),
                                perc_plearn_without_mut = as.numeric(),
                                num_usher = as.numeric(),
                                perc_usher_with_mut = as.numeric(),
                                perc_usher_without_mut = as.numeric(),
                                flag = as.character())

search_vec <- as.vector(searches$lineage)
names(search_vec) <- as.vector(searches$search_term)

for (elem in search_vec) {
  with_mut_defining <- subset(nextclade_all, grepl(names(which(search_vec == elem)), substitutions))
  plearn_with_defining <- nrow(subset(lineages_plearn, taxon %in% with_mut_defining$seqName & lineage == elem))
  plearn_without_defining <- nrow(subset(lineages_plearn, ! taxon %in% with_mut_defining$seqName & lineage == elem))

usher_with_defining <- nrow(subset(lineages_usher, taxon %in% with_mut_defining$seqName & lineage == elem))
usher_without_defining <- nrow(subset(lineages_usher, ! taxon %in% with_mut_defining$seqName & lineage == elem))

perc_plearn_with_mut <- round(100*(plearn_with_defining/nrow(subset(lineages_plearn, lineage                      == elem))), digits = 1)

perc_plearn_without_mut <- round(100*(plearn_without_defining/nrow(subset(lineages_plearn, lineage == elem))), digits = 1)

perc_usher_with_mut <- round(100*(usher_with_defining/nrow(subset(lineages_usher, lineage                      == elem))), digits = 1)

perc_usher_without_mut <- round(100*(usher_without_defining/nrow(subset(lineages_usher, lineage                      == elem))), digits = 1)


to_add <- data.frame(lineage = elem,
                     defining_mut = names(which(ay_vec == elem)),
                     num_plearn = nrow(subset(lineages_plearn, lineage == elem)),
                     perc_plearn_with_mut = perc_plearn_with_mut,
                     perc_plearn_without_mut = perc_plearn_without_mut,
                     num_usher = nrow(subset(lineages_usher, lineage == elem)),
                     perc_usher_with_mut = perc_usher_with_mut,
                     perc_usher_without_mut = perc_usher_without_mut)

if (nrow(subset(lineages_plearn, lineage == elem)) > 0 &
    nrow(subset(lineages_usher, lineage == elem)) > 0 ) {
  
  if (as.numeric(perc_plearn_without_mut) >= 10) {
  flag_plearn <- "plearn discordance over 10%"
} else {
  flag_plearn <- ""
}

if (perc_usher_without_mut >= 10) {
  flag_usher <- "usher discordance over 10%"
} else {
  flag_usher <- ""
}
  to_add$flag <- as.character(paste(flag_plearn, flag_usher, sep="\n"))
  ay_tracking_frame <- rbind(ay_tracking_frame, to_add)
    }

}


datatable(ay_tracking_frame,
           rownames = FALSE,
           escape=FALSE,
          class="compact",
          width = 1000)

# library(kableExtra)
# 
# kable(ay_tracking_frame, format = "html", table.attr = "style='width:30%;'") %>% 
#   kableExtra::kable_styling()

```




